Make any stupid plot you want

### 2021-03-02 draft/sections/abstract.tex
part of #18

To demonstrate the practical value of our model, we propose a model driven re-architecture of the artist layer of the Python visualization library Matplotlib. We can show that using triangulations we can concretely represent the base spaces, make use of programming types for the fiber, and build on Matplotlib's existing infrastructure for the rendering. In addition to providing a way to ensure the library preserves structure, the functional decomposition of the artist in the model could improve modularity, maintainability, and point to ways in which the library provides better support concurrency and interactivity. The thesis will follow through on this proposal to explore how to further develop our model, showing how it can support Matplotlib's current diverse range of data visualizations while providing a better platform for domain-specific visualization library developers.

>>>>

To demonstrate the practical value of our model, we propose a model driven re-architecture of the artist layer of the Python visualization library Matplotlib. We can show that we can concretely represent the base spaces, make use of programming types for the fiber, and build on Matplotlib's existing infrastructure for the rendering. In addition to providing a way to ensure the library preserves structure, the functional decomposition of the artist in the model could improve modularity, maintainability, and point to ways in which the library provides better support concurrency and interactivity. The thesis will follow through on this proposal to explore how to further develop our model, showing how it can support Matplotlib's current diverse range of data visualizations while providing a better platform for domain-specific visualization library developers.

### 2021-03-02 draft/sections/implementation.tex
part of #18

\begin{document}
\section{Prototype Implementation: Matplottoy}
\label{sec:implementation}
\begin{figure}[H]
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/scatter_0.png}
    \end{subfigure}
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/line_1.png}
    \end{subfigure}
    \caption{Scatter plot and line plot implemented using prototype artists and data models, building on Matplotlib rendering.}
    \label{fig:code_scatter_line}
\end{figure}
To prototype our model, we implemented the artist classes for the scatter and line plots shown in figure~\ref{fig:code_scatter_line} because they differ in every attribute: different visual channels \vchannel\ that composite to different marks \vmark\ with different continuities \vindex\.  We make use of the Matplotlib figure and axes artists \cite{hunterArchitectureOpenSource,hunterMatplotlib2DGraphics2007} so that we can initially focus on the data to graphic transformations. 
To generate the images in figure~\ref{fig:code_scatter_line}, we instantiate \mintinline{python}{fig, ax} artists that will contain the new \mintinline{python}{Point, Line} primitive objects we implemented based on our topology model.

@@

\subsection{Artist Class $\vartist^{\prime}$}
\label{sec:code_artist}
The artist is the piece of the matplotlib architecture that constructs an internal representation of the graphic that the render then uses to draw the graphic. In the prototype artist, \mintinline{python}{transform} is a dictionary of the form \mintinline{python}|{parameter:(variable, encoder)}| where parameter is a component in \vfiber, variable is a component in \dfiber,  and the \vchannel\ encoders are passed in as functions or callable objects. The data bundle \dtotal\ is passed in as a \mintinline{python}{data} object. By binding data and transforms to \vartisteq\ inside \mintinline{python}{__init__}, the \mintinline{python}{draw} method is a fully specified artist \vartist.

@@

The components of the visual object, denoted by the Python unpacking convention \mintinline{python}{**visual} are then passed into the \mintinline{python}{assemble} function that is \vmarkd. This assembly function is responsible for generating a representation of the glyph such that it could be serialized to recreate a static version of the graphic. Although \mintinline{python}{assemble} could be implemented outside the class such that it returns an object the artist could then parse to set attributes, the attributes are directly set here to reduce indirection. This artist is not optimized because we prioritized demonstrating the separability of \vchannel\ and \vmarkd. The last step in the artist function is handing itself off to the renderer. The extra \mintinline{python}{*arg, **kwargs} arguments in \mintinline{python}{__init__,draw} are artifacts of how these objects are currently implemnted in Matplotlib.

@@

\begin{figure}[H]
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/bar_v.png}
    \end{subfigure}
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/bar_h.png}
    \end{subfigure}
    \caption{Frequency of Penguin types visualized as discrete bars. }
    \label{fig:code_bar_simple}
\end{figure}
The bar charts in figure~\ref{fig:code_bar_simple} are generated with a \mintinline{python}{Bar} artist. The have the same required \vfiber\ components of (position, length), but in \mintinline{python}{__init__} of \mintinline{python}{Bar} an additional parameter is set, \texttt{orientation} which controls whether the bars are arranged vertically or horizontally. This parameter only applies holistically to the graphic and never to individual data parameters, and highlights how the model encourages explicit differentiation between parameters in \vtotal\ and graphic parameters applied directly to \vmarkd.

@@

        # offset is passed through via assemblers such as multigroup, not supposed to be directly tagged to position 
        position = position + offset

@@

\begin{minted}{python}
@dataclass
class FiberBundle:
"""
Attributes
----------
K: {'tables': []}
F: {variable name: {'type': type, 'moniod', 'range': []}}
"""
    K: dict 
    F: dict
\end{minted}

@@

To generate the scatter plot in figure~\ref{fig:code_scatter_line}, we fully specify a dataset with random keys and values in a section chosen at random form the corresponding fiber component. The fiberbundle \mintinline{python}{FB} is a class level attribute since all instances of \mintinline{python}{VertexSimplex} come from the same fiberbundle.
\begin{minted}{python}
class VertexSimplex: #maybe change name to something else
    """Fiberbundle is consistent across all sections
    """
    FB = FiberBundle({'tables': ['vertex']},  
            {'v1': {'type': float,'monoid':'interval','range': [0,1]},
             'v2': {'type': str, 'monoid':'nominal', 'range':['true', 'false']},
             'v3': {'type': float, 'monoid':'interval', 'range':[2,3]}})

@@

\begin{minted}{python}
class EdgeSimplex: 
    # assign a class level FB attribute
    def __init__(self, num_edges=4, num_samples=1000): 
        self.keys = range(num_edge) #edge id
        # distance along edge
        self.distances = np.linspace(0,1, num_samples)
        # half generlized representation of arcs on a circle
        self.angle_samples = np.linspace(0, 2*np.pi, len(self.keys)+1)

@@

    but differ in assembly \vmarkd. By decomposing the architecture into data, visual encoding, and assembly steps, we are able to build components that are more flexible and also more self contained than the existing code base. While very rough, this API demonstrates that the ideas presented in the math framework are implementable. In choosing a functional approach, if not implementation, we provide a framework for library developers to build reusable encoder \vchannel\, assembly \vmarkd and artists \vartist. We argue that  if these functions are built such that they are equivariant with respect to monoid actions and the graphic topology is a deformation retraction of the data topology, then the artist by definition will be a structure and property preserving map from data to graphic.

>>>>

\begin{document}
\section{Prototype Implementation: Matplottoy}
\label{sec:implementation}
\begin{figure}[H]
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/scatter_0.png}
    \end{subfigure}
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/line_1.png}
    \end{subfigure}
    \caption{Scatter plot and line plot implemented using prototype artists and data models, building on Matplotlib rendering.}
    \label{fig:code_scatter_line}
\end{figure}
To prototype our model, we implemented the artist classes for the scatter and line plots shown in figure~\ref{fig:code_scatter_line} because they differ in every attribute: different visual channels \vchannel\ that composite to different marks \vmark\ with different continuities \vindex\.  We make use of the Matplotlib figure and axes artists \cite{hunterArchitectureOpenSource,hunterMatplotlib2DGraphics2007} so that we can initially focus on the data to graphic transformations. We also exploit the Matplotlib transform stack to transform data coordinates into screen coordinates. To generate the images in figure~\ref{fig:code_scatter_line}, we instantiate \mintinline{python}{fig, ax} artists that will contain the new \mintinline{python}{Point, Line} primitive objects we implemented based on our topology model.

@@

\subsection{Artist Class $\vartist^{\prime}$}
\label{sec:code_artist}
The artist is the piece of the Matplotlib architecture that constructs an internal representation of the graphic that the render then uses to draw the graphic. In the prototype artist, \mintinline{python}{transform} is a dictionary of the form \mintinline{python}|{parameter:(variable, encoder)}| where parameter is a component in \vfiber, variable is a component in \dfiber,  and the \vchannel\ encoders are passed in as functions or callable objects. The data bundle \dtotal\ is passed in as a \mintinline{python}{data} object. By binding data and transforms to \vartisteq\ inside \mintinline{python}{__init__}, the \mintinline{python}{draw} method is a fully specified artist \vartist.

@@

The components of the visual object, denoted by the Python unpacking convention \mintinline{python}{**visual} are then passed into the \mintinline{python}{assemble} function that is \vmarkd. This assembly function is responsible for generating a representation such that it could be serialized to recreate a static version of the graphic. Although \mintinline{python}{assemble} could be implemented outside the class such that it returns an object the artist could then parse to set attributes, the attributes are directly set here to reduce indirection. This artist is not optimized because we prioritized demonstrating the separability of \vchannel\ and \vmarkd. The last step in the artist function is handing itself off to the renderer. The extra \mintinline{python}{*arg, **kwargs} arguments in \mintinline{python}{__init__,draw} are artifacts of how these objects are currently implemnted in Matplotlib.

@@

\begin{figure}[H]
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/bar_v.png}
    \end{subfigure}
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/bar_h.png}
    \end{subfigure}
    \caption{Frequency of Penguin types visualized as discrete bars. }
    \label{fig:code_bar_simple}
\end{figure}
The bar charts in figure~\ref{fig:code_bar_simple} are generated with a \mintinline{python}{Bar} artist. The artist has required visual parameters \vfiber\ of (position, length), and an additional parameter \texttt{orientation} which controls whether the bars are arranged vertically or horizontally. This parameter only applies holistically to the graphic and never to individual data parameters, and highlights how the model encourages explicit differentiation between parameters in \vtotal\ and graphic parameters applied directly to \vmarkd.

@@

        # offset is passed through via assemblers such as multigroup, 
        # not supposed to be directly tagged to position 
        position = position + offset

@@

\begin{minted}{python}
@dataclass
class FiberBundle:
"""
Attributes
----------
K: {'tables': []}
F: {variable name: type}
"""
    K: dict 
    F: dict
\end{minted}

@@

To generate the scatter plot in figure~\ref{fig:code_scatter_line}, we fully specify a dataset with random keys and values in a section chosen at random form the corresponding fiber component. The fiberbundle \mintinline{python}{FB} is a class level attribute since all instances of \mintinline{python}{VertexSimplex} come from the same fiberbundle.
\begin{minted}{python}
class VertexSimplex: #maybe change name to something else
    """Fiberbundle is consistent across all sections
    """
    FB = FiberBundle({'tables': ['vertex']},  
            {'v1': float, 'v2': str, 'v3': float})

@@

\begin{minted}{python}
class EdgeSimplex:

        FB = FiberBundle({'tables': ['vertex','edge']},
                         {'x' : float, 'y':  float, 
                         'color':mtypes.Color()}})
    def __init__(self, num_edges=4, num_samples=1000): 
        self.keys = range(num_edge) #edge id
        # distance along edge
        self.distances = np.linspace(0,1, num_samples)
        # half generlized representation of arcs on a circle
        self.angle_samples = np.linspace(0, 2*np.pi, len(self.keys)+1)

@@

    but differ in assembly \vmarkd. By decomposing the architecture into data, visual encoding, and assembly steps, we are able to build components that are more flexible and also more self contained than the existing code base. While very rough, this API demonstrates that the ideas presented in the math framework are implementable. For example, the \mintinline{python}{draw} function that maps most closely to \vartist\ is functional, with state only being necessary for bookkeeping the many inputs that the function requires. In choosing a functional approach, if not implementation, we provide a framework for library developers to build reusable encoder \vchannel\, assembly \vmarkd and artists \vartist. We argue that  if these functions are built such that they are equivariant with respect to monoid actions and the graphic topology is a deformation retraction of the data topology, then the artist by definition will be a structure and property preserving map from data to graphic.

### 2021-03-02 draft/sections/intro.tex
part of #18

Often this topology has metadata associated with it, describing for example when or where the measurement was taken. Building on the idea of metadata as \textit{keys} and their associated \textit{values} proposed by Munzner \cite{munznerChDataAbstraction}, we propose that information rich metadata are part of the components and instead the values are keyed on coordinate free structural ids. In contrast to Munzner's model where the semantic meaning of the key is tightly coupled to the position of the value in the dataset, our model considers keys to be a pure reference to topology. This allows the metadata to be altered, for example by changing the coordinate systems or time resolution, without imposing new semantics on the underlying structure.

@@

\subsection{Contribution}
This work presents a mathematical model of the transformation from data to graphic representation and a proof of concept implementation. Specifically, the contributions of this work are

\begin{enumerate}
    \item a formal description of the topology preserving relationship between data and graphic via continuous maps
    \item a formal description of the property preservation from data component to visual representation as equivariant maps that carry a homomorphism of monoid actions
    \item abstraction of data as a fiber bundle with schema like fibers to encode components and topology 
    \item algebraic sum operator such that more complex visualizations can be built from simple ones 
    \item a functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model
    \item a prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model
\end{enumerate}

In contrast to mathematical models of visualization that aim to evaluate visualization design, we propose a topological framework for building tools to build visualizations. We defer judgement of expressivity and effectiveness to developers building domain specific tools, but provide them the framework to do so. 
\end{document}

>>>>

Often this topology has metadata associated with it, describing for example dependent variables and the independent variables they are dependent on, or information about the structure of the data such as when and where the measurement was taken. Building on the idea of metadata as \textit{keys} and their associated \textit{values} proposed by Munzner \cite{munznerChDataAbstraction}, we propose that information rich metadata are part of the components and instead the values are keyed on coordinate free structural ids. In contrast to Munzner's model where the semantic meaning of the key is tightly coupled to the position of the value in the dataset, our model considers keys to be a pure reference to topology. This allows the metadata to be altered, for example by changing the coordinate systems or time resolution, without imposing new semantics on the underlying structure.

@@

\subsection{Contribution}
This work presents a mathematical model of the transformation from data to graphic representation and a proof of concept implementation. Specifically, the contributions of this work are 
\begin{enumerate}
    \item a formal description of the topology preserving relationship between data and graphic via continuous maps
    \item a formal description of the property preservation from data component to visual representation as equivariant maps that carry a homomorphism of monoid actions
    \item abstraction of data structure using fiber bundles with schema like fibers to encode components and topology 
    \item algebraic sum operator such that more complex visualizations can be built from simple ones 
    \item a functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model
    \item a prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model
\end{enumerate}
In contrast to mathematical models of visualization that aim to evaluate visualization design, we propose a topological framework for building tools to build visualizations. We defer judgement of expressivity and effectiveness to developers building domain specific tools, but provide them the framework to do so. 
\end{document}
